---
title: Data cleaning
author: Mike O'Brien
date: last-modified
execute:
  message: false
  warning: false
---

```{r}
library(arrow)
library(dplyr)
library(ggplot2)
```

# Striped bass
```{r}
#| label: sb-import
sb <- open_dataset("../data/RAW/combined_striped_bass.parquet") |>
  filter(longitude != 0, !is.na(datetime)) |> # CHANGE THIS LATER!
  mutate(latitude = abs(latitude), wk = week(datetime)) |>
  group_by(transmitter, year, wk, group) |>
  summarize(
    med_lon = median(longitude, na.rm = TRUE),
    med_lat = median(latitude, na.rm = TRUE)
  ) |>
  collect()
```

First visualize the distribution of the median per-individual latitudes through time:

```{r}
#| label: sb-plot
#| fig-cap: "Weekly median latitude of striped bass. Colors represent putative populations; DE: Delaware, HR: Hudson River, PR: Potomac River."
ggplot(sb) +
  geom_point(aes(x = wk, y = med_lat, color = group), alpha = 0.4) +
  facet_wrap(~year) +
  labs(x = "Week of year", y = "Weekly median latitude") +
  theme_minimal() +
  theme(axis.title = element_text(size = 12))
```

It seems that we have resident striped bass from both the Delaware and Potomac River. Striped bass exhibit partial migration, where fish <80 cm remain resident in their natal estuary [@secor2020] (though our understanding of this is rapidly changing!). The projects providing the data for Delaware and Potomac fish purposefully tagged smaller, resident fish. As a first pass, we'll remove any fish that didn't move higher than 40.46 $\degree$ latitude (the latitude of Trenton, NJ).

```{r}
#| label: sb-import-migrants
#| echo: false
drop <- open_dataset("../data/RAW/combined_striped_bass.parquet") |>
  filter(longitude != 0, !is.na(datetime)) |>
  group_by(transmitter, year) |>
  summarize(max_lat = max(latitude, na.rm = TRUE)) |>
  filter(max_lat < 40.26) |>
  select(transmitter, year) |>
  collect()

sb <- open_dataset("../data/RAW/combined_striped_bass.parquet") |>
  anti_join(drop) |>
  mutate(latitude = abs(latitude), wk = week(datetime)) |>
  group_by(transmitter, year, wk, group) |>
  summarize(
    med_lon = median(longitude, na.rm = TRUE),
    med_lat = median(latitude, na.rm = TRUE)
  ) |>
  collect()


ggplot(sb) +
  geom_point(aes(x = wk, y = med_lat, color = group), alpha = 0.4) +
  facet_wrap(~year) +
  labs(x = "Week of year", y = "Weekly median latitude") +
  theme_minimal() +
  theme(axis.title = element_text(size = 12))
```

Great, now we definitely have some migratory striped bass.


```{r}
#| eval: false
open_dataset("data/RAW/combined_striped_bass.parquet") |>
  anti_join(drop) |>
  mutate(latitude = abs(latitude), wk = week(datetime)) |>
  write_parquet("data/derived/striped_bass.parquet")
```