---
title: Geodesic Regression
format: html
---

::: {.callout-note collapse="true"}
Pardon our dust: Narrative of this section is poor and mostly an outline to ourselves.

- Updated on May 21, 2026 to include all available striped bass data.
:::


## Geoid and Geodesics
  - Explain geoid
    - Imagine circle through the center of the Earth with the Earth's radius
    - Line from N pole to S pole through Greenwich, along the 0* meridian
    - Continue from S pole to N pole along the 180* meridian
    - Can transform this by changing the meridian (longitude), changing the poles (latitude), or both
  - Explain geodesic
    - Line along the surface of this circle

## Geodesic regression
  - Even here, we are projecting into Euclidean space to make math simpler
  - Rienmannian manifolds -- a way to express curvature

## As it applies to ROSA project

### Geodesic measure
  - Convert lat/lon coordinates to unit sphere (S2 library)
    - Assumptions and innacuracies here, however:
        - spheres are well-defined Reinmanninan manifolds
        - I have not found geoids (WGS84/GRS80) defined as Reinmanninan manifolds. I am unsure if this is because people have already tried it and it's a dead end or if the math is just new.
  - Conduct geodesic regression to find best-fit geodesic through points

### Comparisons
  - Compare this line against least-squares fit in common "projections": unprojected, Mercator, UTM (17-19), AEQD, gnomonic

```{r}
#| echo: false
if (grepl("notebooks", getwd())) {
  Sys.setenv(DATA_DIR = "../")
} else {
  Sys.setenv(DATA_DIR = "")
}
```

# Try crossing R/python

Set up python, make sure `geomstats` is installed
```{r}
library(reticulate)
py_require("geomstats")
```

Import data

```{r}
library(arrow)
library(dplyr)
library(sf)
library(s2)
library(ggplot2)

sb_file <- file.path(
  Sys.getenv("DATA_DIR"),
  "data/derived/striped_bass.parquet"
)
sb <- open_dataset(sb_file) |>
  group_by(transmitter, year, wk, group) |>
  filter(latitude > 34) |> # detections almost definitely wrong and needs to be removed
  summarize(
    mean_lon = mean(longitude, na.rm = TRUE),
    mean_lat = mean(latitude, na.rm = TRUE)
  ) |>
  collect()

# Viz in orthographic (globe) projection
sb |>
  st_as_sf(coords = c("mean_lon", "mean_lat"), crs = 4326) |>
  ggplot() +
  geom_sf() +
  geom_sf(data = st_sfc(st_point(c(-73.04283, 40.32766)), crs = 4326), ) +
  coord_sf(crs = '+proj=ortho +lon_0=-75 +lat_0=32')
```

The [Geomstats Hypersphere module](https://geomstats.github.io/api/geomstats.geometry.html#module-geomstats.geometry.hypersphere) prefers coordinates on the unit circle: X, Y, and Z values between -1 and 1. However, latitude and longitude coordinates are angles. So, we need to convert coordinates to the unit cirlce.

The [s2 R package](https://r-spatial.github.io/s2/) interfaces with the [S2 C++ library](https://s2geometry.io/). One of the shortcuts this library takes is to approximate the Earth as a unit sphere, which means it has a nice built-in function to convert coordinates to the unit sphere. This allows it to avoid a bunch of computationally-intensive trigonometric functions. I'll convert the mean striped bass locations above to the unit sphere. 
```{r}
sb_s2 <- s2_lnglat(sb$mean_lon, sb$mean_lat) |>
  as_s2_point() |>
  # Geomstats package works best with matrices/arrays
  as.matrix()
```

First, import the Hypersphere submodule, create a 2 dimensional sphere, then confirm that the points above lie on that sphere.
```{r}
gs <- import("geomstats.backend")
hypersphere <- import("geomstats.geometry.hypersphere")$Hypersphere


# need to use as.integer or reticulate will pass a float, which Hypersphere does not like
sphere <- hypersphere(dim = as.integer(2))

sb_s2 |>
  sphere$belongs() |>
  all()
```

To run a tangent PCA, which I currently believe is equivalent to a geodesic PCA but could be incorrect, we need to find a point on the sphere that the tangent plane will touch. The [Geomstats walkthrough](https://geomstats.github.io/notebooks/06_practical_methods__riemannian_frechet_mean_and_tangent_pca.html) uses the [Frechet mean](https://en.wikipedia.org/wiki/Fr%C3%A9chet_mean).

```{r}
frechet_mean <- import("geomstats.learning.frechet_mean")$FrechetMean

mean_fun <- frechet_mean(sphere)
mean_fit <- mean_fun$fit(sb_s2)
mean_estimate <- mean_fit$estimate_
```

This corresponds to a lon/lat of:

```{r}
frechet_mean_coord <- s2_point(
  mean_estimate[1],
  mean_estimate[2],
  mean_estimate[3]
) |>
  as_s2_lnglat()

frechet_mean_coord
```

Whereas averaging the lon/lat coordinates is:

```{r}
raw_mean <- s2_lnglat(mean(sb$mean_lon), mean(sb$mean_lat))
raw_mean
```

How far close is this to the point returned by averaging the lon/lat coordinates?

```{r}
location_distance <- s2_distance(frechet_mean_coord, raw_mean)
location_distance
```

About `r round(location_distance / 1000, 1)` kilometers away.

Now we'll run a PCA on a tangent plane at this point.

```{r}
tpca <- import("geomstats.learning.pca")$TangentPCA

tpca_mod <- tpca(sphere, n_components = as.integer(2))
tpca_mod$fit(sb_s2, base_point = mean_estimate)

# transform data onto tangent plane
tangent_sb <- tpca_mod$transform(sb_s2)
plot(tangent_sb)
```

How much variation is explained by this PCA?

```{r}
exp_var <- tpca_mod$explained_variance_ratio_
barplot(c(PC1 = exp_var[1], PC2 = exp_var[2]))
```
The first principal component explains `r round(exp_var[1] * 100, 1)`% of the variance, with the second principal component explaining the remaining `r round(exp_var[2] * 100), 1)`%.

Now some data manipulation to visualize the geodesic PCA:

```{r}
geodesic_1 = sphere$metric$geodesic(
  initial_point = mean_estimate,
  initial_tangent_vec = tpca_mod$components_[1, ] |> t()
)
geodesic_2 = sphere$metric$geodesic(
  initial_point = mean_estimate,
  initial_tangent_vec = tpca_mod$components_[2, ] |> t()
)

# for whatever reason I can't make this work with R's `seq`, so using the geomstats version
geodesic_1_pts <- geodesic_1(gs$linspace(-0.5, 0.5, as.integer(100))) |>
  # creates a 1*100*3 matrix that should be 100*3. Drop removes the empty dimension, giving 100*3
  drop()
g1_s2 <- s2_point(
  geodesic_1_pts[, 1],
  geodesic_1_pts[, 2],
  geodesic_1_pts[, 3]
) |>
  as_s2_lnglat()
geodesic_2_pts <- geodesic_2(gs$linspace(
  as.integer(-1),
  as.integer(1),
  as.integer(100)
)) |>
  drop()
g2_s2 <- s2_point(
  geodesic_2_pts[, 1],
  geodesic_2_pts[, 2],
  geodesic_2_pts[, 3]
) |>
  as_s2_lnglat()
```

These are XYZ on the unit circle. Use S2 to convert them to lat/lon and visualize.

```{r}
plot(s2_lnglat(sb$mean_lon, sb$mean_lat), col = "black")
plot(g1_s2, add = T, type = 'l', col = 'red')
plot(g2_s2, add = T, type = 'l', col = 'orange')
```

```{r}
aeqd_proj <- paste0(
  '+proj=aeqd +lon_0=',
  s2_x(frechet_mean_coord),
  ' +lat_0=',
  s2_y(frechet_mean_coord)
)
ggplot() +
  geom_sf(
    data = st_as_sf(sb, coords = c("mean_lon", "mean_lat"), crs = 4326),
    color = 'blue'
  ) +
  geom_sf(data = st_as_sf(g1_s2)) +
  geom_sf(data = st_as_sf(g2_s2)) +
  coord_sf(
    xlim = c(-4e5, 3.5e5),
    ylim = c(-4e5, 3.5e5),
    crs = aeqd_proj
    #OMERC
    # crs = paste0('+proj=omerc +gamma=-39 +lonc=',s2_x(frechet_mean_coord), ' +lat_0=', s2_y(frechet_mean_coord))
  )
```


# Try AEQD then PCA

Project through the Frechet mean.
```{r}
sb_sf <- st_as_sf(
  sb,
  coords = c("mean_lon", "mean_lat"),
  crs = 4326
) |>
  st_transform(aeqd_proj)

pca <- prcomp(st_coordinates(sb_sf))
rng <- predict(pca, st_coordinates(sb_sf))[2, ] |> range()

pca1_sf <- cbind(
  seq(-1e6, 1e6, length.out = 100),
  rep(0, 100)
) %*%
  t(pca$rotation) |>
  as.data.frame() |>
  st_as_sf(
    coords = c("X", "Y"),
    crs = aeqd_proj
  ) |>
  st_combine() |>
  st_cast("LINESTRING")

gamma <- do.call(atan2, c(1e3, 0) %*% t(pca$rotation) |> as.list()) * 180 / pi
omerc_proj <- paste0(
  '+proj=omerc +gamma=',
  gamma,
  ' +lonc=',
  s2_x(frechet_mean_coord),
  ' +lat_0=',
  s2_y(frechet_mean_coord)
)

rot_plot <- ggplot() +
  geom_sf(data = pca1_sf) +
  geom_sf(data = sb_sf) +
  geom_sf(
    data = st_as_sf(g1_s2) |>
      st_combine() |>
      st_cast("LINESTRING") |>
      st_transform(aeqd_proj),
    col = 'red'
  ) +
  geom_sf(
    data = (c(0, 0) %*%
      t(pca$rotation) |>
      st_point() |>
      st_sfc(crs = aeqd_proj)),
    col = 'blue',
    size = 5
  ) +
  coord_sf(
    xlim = c(-2.5e5, 1.5e5),
    ylim = c(-5e5, 5e5),
    crs = '+proj=omerc +gamma=-43.4061 +lonc=-73.04283 +lat_0=40.32766'
  ) +
  theme(
    axis.text = element_text(size = 14, angle = 45),
    axis.text.x = element_text(hjust = 1, angle = 0)
  )

rot_plot
```



```{r}
ragg::agg_png(
  height = 6,
  width = 3,
  units = "in",
  filename = "rotation_plot.png",
  res = 300
)
rot_plot
dev.off()
```