---
title: Projection Error
date: 2026-03-23
format: html
---

```{r}
# remotes::install_github("hypertidy/geographiclib")
library(geographiclib)
library(data.table)
library(sf)
library(ggplot2)
library(patchwork)
sb <- qs2::qs_read("data/RAW/hud_detects.qs2") |>
  as.data.table() |> 
  unique(by = c("long", "lat")) |> 
  _[, .(lon = long, lat = lat)]
atl <- fread("data/RAW/fixed_rounded_station_locations.csv") |> 
  unique(by = c("r.lon", "r.lat")) |> 
  _[, .(lon = r.lon, lat = r.lat)]
st <- fread("data/RAW/sandtigers.csv") |> 
  unique(by = c("Longitude", "Latitude")) |> 
  _[, .(lon = Longitude, lat = Latitude)]

locations <- list(sb = sb, atl = atl, st = st)
```

- import striped bass data
- use geographiclib distance calculations between points and migration vector as "truth"
- project into various reference systems
  - AEQD
  - UTM
    - 18
    - 19
    - Note that sturgeon and sharks will go to 17
  - Oblique Mercator
  - Oblique Cassini
- calculate distance metrics
  - pointwise
    - absolute error
    - squared error
  - projection wise
    - MAE
    - RMSE

```{r}
truth <- lapply(locations, geodesic_distance_matrix)
centers <- lapply(
  locations,
  \(.) .[, .(med_lon = median(lon, na.rm = T), med_lat = median(lat, na.rm = T))]
  )
```

```{r}
proj_dist <- function(stations, centers){
  list(
    aeqd = azeq_fwd(stations, centers[[1]], centers[[2]]) |> 
      dist() |> 
      as.matrix(),
    utm_auto = utmups_fwd(stations)[, 1:2] |>
      dist() |>
      as.matrix(),
    utm17 = tm_exact_fwd(stations, -81) |>
      dist() |>
      as.matrix(),
    utm18 = tm_exact_fwd(stations, -75) |>
      dist() |>
      as.matrix(),
    utm19 = tm_exact_fwd(stations, -69) |>
      dist() |>
      as.matrix(),
    tm = tm_exact_fwd(stations, centers[[1]]) |>
      dist() |>
      as.matrix(),
    cass = cassini_fwd(stations, centers[[1]], centers[[2]])[, 1:2] |>
      dist() |>
      as.matrix()
  )
}

projected_distances <- mapply(proj_dist, locations, centers, SIMPLIFY = F)
```


```{r}
proj_errs <- function(dists, truth, species){
  lapply(dists[[species]],
  \(.) (. - truth[[species]])[lower.tri(truth[[species]])]
  )
}


errors <- sapply(
  names(projected_distances),
  \(.) proj_errs(projected_distances, truth, species = .),
  simplify = FALSE,
  USE.NAMES = TRUE
) |> 
  lapply(\(.) as.data.table(.) |> melt()) |> 
  rbindlist(idcol = 'species')
```



```{r}
ggplot() +
  geom_histogram(
    data = errors[variable != "utm_auto"],
    aes(x = abs(value), fill = variable),
    position = 'dodge',
    binwidth = 50
  ) +
  geom_vline(xintercept = 500) +
  facet_wrap(~ species, scales = "free", ncol = 1)
```

Number greater than 500m distance (receiver detection range)
```{r}
errors[value > 500, .N, c('variable', 'species')]
errors[, sqrt(mean(value^2, na.rm = T)),  c('variable', 'species')]
errors[, .(min = min(value, na.rm =T),max = max(value, na.rm =T)),  c('variable', 'species')]
```


<!-- RESTART GENERALIZNG CODE OVER SPECIES HERE -->
```{r}
sb_sf<- sb |> 
  as.data.frame() |> 
  st_as_sf(coords = c("mean_lon", "mean_lat"), crs = 4326)
sb_sf$row = 1:nrow(sb_sf)
setDT(sb_sf)
```
```{r}
err_mat <- (tm_dists - truth)
err_df <- data.table(
  row = rownames(err_mat)[row(err_mat)],
  col = colnames(err_mat)[col(err_mat)],
  dist = c(err_mat)
) |>
  _[, let(
    from = pmin(row, col),
    to = pmax(row, col)
  )] |>
  unique(by = c('from', 'to')) |>
  _[, row := as.integer(row)] |>
  _[dist != 0, sqrt(mean(dist^2)), row] |>
  _[sb_sf, on = 'row']

a <- ggplot(data = err_df) +
  geom_sf(aes(geometry = geometry, color = V1)) +
  scale_color_viridis_c() +
  labs(color = 'RMSE (m)', title = "TM (-73 lon_0)") +
  theme(
    legend.position = c(0.75, 0.25)
  )
b <- ggplot(data = err_df) +
  geom_histogram(aes(x = V1)) +
  scale_color_viridis_c() +
  labs(x = "RMSE (m)")

a + b
```